---
title: 'Equivalence scales for household data in RLMS'
author: ''
date: '2022-11-09'
slug: household-equivalence-scales
categories: []
tags: []
---


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}


library(data.table)
library(tidyverse)
library(viridis)
library(gt)

load(file = "C:\\Users\\Ivan\\Documents\\Projects\\Blog\\Sandbox\\ru.Rdata")

# rename measures ---------------------------------------------------------


setnames(ru, 'ID_H', 'hhid') 
setnames(ru, 'idind', 'pid') 
setnames(ru, 'I1', 'born') 
setnames(ru, 'I2', 'cob')
setnames(ru, 'J1', 'employed')
setnames(ru, 'J10', 'wages')
setnames(ru, 'J60', 'income')
setnames(ru, 'H5', 'gender')




# correct for extreme values ----------------------------------------------




ru <-  ru %>% 
  mutate(income = ifelse(income > 99999994, NA, income))

ru[, income_denom := ifelse(year < 1998, income/1000, income)]

ru[, wage_denom := ifelse(year < 1998, wages/1000, wages)]

# sort --------------------------------------------------------------------
setcolorder(ru, c("hhid", "year", "pid"))

setorder(ru, hhid, year, pid)



# equivalence scale B -----------------------------------------------------

ru[, n_within_hhid := rowid(hhid, year)]


ru[, equiv := case_when(n_within_hhid == 1 ~ 1,
                        n_within_hhid > 1 & age > 18 ~ 0.5,
                        n_within_hhid > 1 & age < 18 ~ 0.3)]

ru[,  hh_equiv := sum(equiv) , by = .(hhid, year)]
ru[,  hh_income := sum(income_denom, na.rm=T) , by = .(hhid, year)]
ru[,  hh_equiv_income := hh_income/hh_equiv , by = .(hhid, year)]


```




# Introduction 


In the last few posts, I listed measures of inequality using individual income, without adjusting for household size. 
Other authors, especially authors who write about poverty and deprivation,  typically adjust for household size by considering a weight for total household income. These measures are usually presented in the data, but the RLMS does not consider them explicitly. 

Equivalence scales help us assign a weight of “need” to different household types and to their total household income. 
Authors typically consider three measures for calculating equivalence scales. 

1. The size of the household 

2. The age of its members (whether they are adults or children) 

3. If using longitudinal data, the year that the data was collected

In this post, I compare two scale the original "OECD scale", and the "OECD-modified scale".
The OECD scale assigns a value of 1 to the first household member, of 0.7 to each additional adult and of 0.5 to each child. 
The OECD Modified scale assigns a value of 1 to the household head, of 0.5 to each additional adult member and of 0.3 to each child. 
Other scales exist, including scales developed by the OECD, but we ignore them here. 

## A brief example

Focusing on the modified scale, I show the basic shape of the equivalence scale and the household’s total equivalence weight below. 
In this example, we look at just one household (hhid=10013) over several years (from 2010 to 2014 in this example). 
Here, the first person is a 33 year old man with a personal ID marked 1001301. 
This person gets a value of 1. 
Every subsequent person aged over 18 in the household gets a value of 0.5, in this case this is the younger woman (pid 25026). 
Lastly, there is a young 3 year old child in the home, and they receive a value of 0.3.  
This brings the total equivalised weight to the household of 1.8 (1.0 + 0.5 + 0.3 = 1.8) for each household member, and each year.

```{r, echo=FALSE}


t1<- ru[hhid==10013 & year>2009, .(hhid, year, pid, gender, age, income, equiv, hh_equiv)]

gt(t1) %>% 
  tab_header(
    title = md("**One household over time**"),
  subtitle = "Following one household 2010-2020") %>% 
  tab_source_note(source_note = "Source: RLMS analysis") %>% 
  tab_spanner(label = "Identifiers",
    columns = c(hhid, year, pid)) %>%
  tab_spanner(label = "Demographics",
    columns = c(gender, age, income)) %>% 
  tab_spanner(label = "Equivalence scale",
    columns = c(equiv, hh_equiv)) %>% 
  fmt_number(
    columns = c(income),
    decimals = 0,
    use_seps = TRUE)

```


If we consider this effect on income, we get the following data frame. 
Again, I am looking at one household over time (10013). 
In 2010, this household had three people living in it, and their equivalence scale added up to 1.8.
Their total income came from the 33 year old male in 2010, who earned 22,000 rubles per month, and so their equivalised household income came to 12,222 rubles per month. 
In 2011, this household saw a change in their income. 
Although the man’s income remained the same (22,000) the woman’s income increased to 8,200 rubles per month. 
As a result, their equivalised household income increased to 16,778 rubles per month.
In short we are showing that the total household income is divided 1.8 ways. 


```{r, echo=FALSE}


t2 <- ru[hhid==10013 & year>2009, .(hhid, year, pid, gender, age,  hh_equiv, income, hh_equiv_income)]

gt(t2) %>% 
  tab_header(
    title = md("**One household over time**"),
  subtitle = "Following one household 2010-2020") %>% 
  tab_source_note(source_note = "Source: RLMS analysis") %>% 
  tab_spanner(label = "Identifiers",
    columns = c(hhid, year, pid)) %>%
  tab_spanner(label = "Demographics",
    columns = c(gender, age)) %>% 
  tab_spanner(label = "Income",
    columns = c(hh_equiv, income, hh_equiv_income)) %>% 
  fmt_number(
    columns = c(income, hh_equiv_income),
    decimals = 0,
    use_seps = TRUE)


```

Moving forward with this example (the OECD modified scale), I show how the scale affects total household income, and compare it to individual level income.
In previous posts, I would have focused on mean income but people tend to live in households which pool their income together, as shown by the trend in mean HH income.
First, all of the three eates all rose over time. 
Second, mean individual income is lower than total household income, for obvious reasons, and equivalised household income sits between these two measures as it adjusts for household size. 
In total equivalised household income sits closer to individual measures of income than the household total. 


```{r, echo=FALSE}

load(file = "C:\\Users\\Ivan\\Documents\\Projects\\Blog\\Sandbox\\p_equiv")

p_equiv


```



## Comparing two scales

Finally, I want to briefly compare the two OECD measures of equivalisation. 
How do they impact total household income?
We see that the modified scales assumes a higher average amount of income, when comapred to the original scale. In this way the modified scale moves closer to the total and unadjusted household amount of income. 
Which scale to use will have important implications for measures of poverty and income inequality, but it seems that most research have picked up the modified scale. 


```{r, echo=FALSE}

load(file = "C:\\Users\\Ivan\\Documents\\Projects\\Blog\\Sandbox\\comp")

comp

```