---
title: Parents' Expectations 2004-2018
author: Privlko
date: '2018-07-02'
slug: parents-expectations-2004-2018
categories:
  - Levada
  - Occupations
  - Labour Market
tags:
  - Levada
  - Labour Markets
---


A recent post from Russia's [Levada Centre](https://www.levada.ru/2018/06/21/shkolnye-predmety-i-professii/) caught my attention. On the topic of school and future careers, parents are asked what they would like their children to do, what professions they prefer them to choose. 

Levada collected this data from 2004, so it's possible to track parents' expectations of their kids over time (somewhat).

The format is not ideal, occuaptions seem to be in broad categories, not all of which are related. For example, sociology, political science etc are all lumped together, but likely reflect the low interest in each of these occupations. However, it's possible to catch interesting changes between 2004 and 2018.



```{r, include=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(ggthemes)
library(lubridate)
library(ggplot2)


school <- read_csv("https://raw.githubusercontent.com/Privlko/hello-world/master/levada.csv",
                   skip = 2) 

school <- school %>%
  gather('may.04', 'jun.06' ,
         'jun.07', 'jul.10',
         'may.14', 'may.15',
         'may.16', 'may.18', key = "mnth", value = "portion")



school$mnth <- parse_date(school$mnth, "%b.%y")
school$portion <- parse_vector(school$portion, col_integer(), na = c("<1"))
school$Occupation <- parse_factor(school$Occupation, levels = rev(unique(school$Occupation)))
school$point <- as.factor(school$mnth)


b <- school %>%
  group_by(Occupation, point) %>%
  filter(point == "2004-05-01" | point == "2018-05-01") %>%
  summarise(portion = sum(portion, na.rm = TRUE)) %>%
  arrange(desc(portion))


b

lab_right <- 
  b %>%
  group_by(Occupation) %>%
  arrange(desc(portion)) %>%
  top_n(1)


lab_left <- b %>%
  group_by(Occupation) %>%
  arrange(desc(portion)) %>%
  slice(2)

big_diff <- b %>% 
  spread(point, portion) %>% 
  group_by(Occupation) %>% 
  mutate(Max = max(`2004-05-01`, `2018-05-01`),
         Min = min(`2004-05-01`, `2018-05-01`),
         Diff = Max / Min - 1) %>% 
  arrange(desc(Diff)) %>%
  filter(Diff > .2)

lab_right <- filter(lab_right, Occupation %in% big_diff$Occupation)
lab_left <- filter(lab_left, Occupation %in% big_diff$Occupation)

highlight <- filter(b, Occupation %in% big_diff$Occupation)


p <-  ggplot(b, aes(portion, Occupation)) +
  geom_line(aes(group = Occupation),
            alpha=0.2)+
  geom_point(aes(color = point), size = 2.5) +
  geom_text(data = lab_right, 
            aes(colour=point, label=round(portion,0)),  
            size=3, hjust = -1.5) +
  geom_text(data = lab_left, 
            aes(colour=point, label=round(portion,0)),  
            size=3, hjust = 2) +
  scale_x_continuous(limits = c(-5, +30))


my_title <- "Parents have become less interested in kids \nbecoming lawyers, and more interested in \nkids becoming soldiers"
my_subtitle <- "When asking parents, which profession they would like for their children, \nlaw, medicine, and business remain the top industries. \nSupport for military careers has also grown."
my_caption <- "Source: Levada Centre \nPlot: @privlko"

q1 <- p + 
  scale_color_discrete(labels = c("2004", "2018")) +
  scale_y_discrete(expand = c(.02, 0))+
  labs(title= my_title,
       subtitle= my_subtitle,
       caption=my_caption,
       x = "Portion of parents choosing the occupation",
       y = "Occupations and profession") +
  theme_minimal()+
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 17, margin = margin(b = 10)),
        plot.subtitle = element_text(size = 10, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 8, margin = margin(t = 10), color = "grey70", hjust = 0),
        axis.title.y = element_text(size = 8, margin = margin(t = 10), color = "darkslategrey", hjust = 0),
        axis.title.x = element_text(size = 8, margin = margin(t = 10), color = "darkslategrey", hjust = 0),
        axis.text.y = element_text(size = 8, margin = margin(t = 10), color = "darkslategrey", hjust = 1)
  )


```

The resulting graph looks like this.

```{r, echo=FALSE,warning=FALSE}
q1
```


A number of patterns emerge. The first is the sharp fall in interest in the legal profession. In 2004, the law was one of the most sought after occupations. By 2018, interest fell significantly. Interest has alsoo fallen in common trades. Instead, parents have become more interested in children taking military careers, and political positions. Some interest has also grown in teaching and elementary occupations, although these remain unpopular overall. 

There is some growth in the medical profession, but this category has remained a top favourite in Russia, which likely reflect other countries also. 

The growth of interest in military careers, could be cultural, but it could also reflect changes in pay and conditions in this sector. 

Levada is a great source of public opinion polling, and I'm looking forward to exploring this data more. 

