---
title: Public Thoughts on Politics and Government
author: Ivan Privalko
date: '2018-01-28'
slug: public-thoughts-on-politics-and-government
categories:
  - politics
  - ESS
  - sociology
tags: []
---

The European Social Survey for 2016 lists variables on government and politics. These are worth exploring, especially with the rise of protests and public meetings against the Kremlin. Often, critics claim these are disingenuous movements, made to create media content in Europe and the States. Typically, they point to Putin's high approval rating as a symbol that all is well, and the population holds pro-government views. 

In this post I explore some specifics, how do Russians feel about government, their ability to impact policy, and their ability to influence political outcomes?

Is public opinion, overall, really pro-government, or does the public hold a pro-Putin view despite dissatisfaction in other aspects of politics? I explore these below while trying out the [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/).

```{r, include=FALSE}
library(readstata13)
library(haven)
library(tidyverse)
library(ggplot2)

stata.data <- read_dta("C:/Users/Administrator.admin-PC2/Desktop/Directory/Blog Ideas/ESS8RU.stata/ESS8RU.dta")

```

As a side note (mostly to myself), the first obstacle is loading the data. Currently, ESS does not provide data in R format. Luckily, several packages can convert Stata data files into R format. I'm using 

> library(haven)

However others claim readstata13 works well too 

> library(readstata13)

Both reformat Stata data into R. I limit the file to a few demographics variables, and several measures of government satisfaction. The relevant code is listed below.


```{r}
poli <- stata.data %>%
  select(hincfel, gndr, polintr, psppsgva, psppipla, stfgov, gvrfgap)
```


```{r, include=FALSE}

# recode factors-------------------------------------------------------------------
poli$interest = factor(poli$polintr)
poli$interest = factor(poli$interest,labels=c("Very interested","interested","hardly", "not at all", "refused", "don't know"))

poli$interest[poli$interest=="refused"] <- NA
poli$interest[poli$interest=="don't know"] <- NA

###gender
poli$gndr = factor(poli$gndr)
poli$gndr = factor(poli$gndr,labels=c("Male","Female"))


#allows to have say

poli$psppsgva = factor(poli$psppsgva)
poli$psppsgva = factor(poli$psppsgva,labels=c("Not at all","very little", "some", "a lot", "a great deal", NA, NA))


#allows to influence


poli$psppipla = factor(poli$psppipla)
poli$psppipla = factor(poli$psppipla,labels=c("Not at all","very little", "some", "a lot", "a great deal", NA, NA))

##satisfaction with gov
poli$stfgov[poli$stfgov>60] <- NA

##feelings about household income
poli$hincfel = factor(poli$hincfel)
poli$hincfel = factor(poli$hincfel,labels=c("comfortable","coping", "difficult", "very difficult", NA, NA))

```

The frame above can be laid out in a few bar charts. Most of all I'm interested in whether Russians are interested in politics, whether they feel they can influence politics, and their satisfaction with the government. I also feel like there may be a difference in opinion between people who are doing well, and those who are struggling financially. I explore this in some of the output. 

#Interest in politics

Firstly, it's possible that the majority is not interested in politics. Those who are willing to go out and protest may be part of a unique group, one that is particularly interested in the political process. It is worth asking, are people interested in politics overall?

```{r, include=FALSE}

# Very basic bar graph
p <- ggplot(data=poli, aes(x=interest)) 

```

```{r, echo=FALSE}
p + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Political interest",
         subtitle = "Factor variable measuring interest",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) 

```

Interest appears to be split, with half of respondents either interested or disinterested in politics. It is worth checking how opinion differs by feelings about income. 

```{r, include=FALSE}

# Very basic bar graph
q <- ggplot(data=poli, aes(x=hincfel, fill=interest)) 

```

```{r, echo=FALSE, warning=FALSE}

q + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Political interest by feelings about income",
         subtitle = "Factor variable measuring interest",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) +
  xlim("comfortable", "coping","difficult","very difficult")

```

The data suggests that feelings about household income do not have an impact on interest in politics. In each group, respondents are either interested in politics or hardly interested. Overall, Russians appear split on the topic. 

#Government allows you to have a say

How do people feel about their ability to voice their opinion? Do respondents feel like they have a say on political outcomes? Below, the chart shows who is likely to say that the government allows them to have a say in the political process. 

```{r, include=FALSE}

# Very basic bar graph
p <- ggplot(data=poli, aes(x=psppsgva)) 

```

```{r, echo=FALSE, warning=FALSE}
p + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Government allows to have a say",
         subtitle = "Factor variable measuring opinion",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) 

```

Here, the majority do not accept that they are allowed have a say in the political process. Although 500+ respondents believe they may participate "somewhat", most respondents do not accept that they have a say in the political process. Here, it is also worth asking if struggling households stand out.

```{r, include=FALSE}

# Very basic bar graph
q <- ggplot(data=poli, aes(x=hincfel, fill=psppsgva))

```

```{r, echo=FALSE, warning=FALSE}
q + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Government allows to have a say by household income",
         subtitle = "Factor variable measuring opinion",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) +
  xlim("comfortable", "coping","difficult","very difficult")

```


Each income group believes they have no voice in the political process. However, respondents who feel they are "comfortable" are somewhat likely to say they have "some" influence. However, the distribution shows this is a small group of people. The remaining groups feel they have no "voice" on the political process. Respondents who find it "very difficult" to get by of their income seem like the most likely to have "no voice".

Having political voice is one thing, but a better measure may lie in whether they can influence an election. People may hold dissenting views, and feel like they have no voice, but they may also feel like their vote or influence can change a political outcome. One question asks respondents how they feel about their influence on politics. 


```{r, include=FALSE}

# Very basic bar graph
p <- ggplot(data=poli, aes(x=psppipla)) 

```

```{r, echo=FALSE, warning=FALSE}
p + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Government allows to have influence on politics",
         subtitle = "Factor variable measuring opinion",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) 

```


Here too, most respondents feel as though they have no influence on the political process or on political outcomes. As with the previous measure there may be a crude class effect. Those who are "comfortable" may feel like they have a greater impact on politics than those who struggle to get by.

```{r, include=FALSE}

# Very basic bar graph
q <- ggplot(data=poli, aes(x=hincfel, fill=psppipla))

```

```{r, echo=FALSE, warning=FALSE}
q + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Government allows to have influence, by household income",
         subtitle = "Factor variable measuring opinion",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) +
  xlim("comfortable", "coping","difficult","very difficult")
```

Each income group feels like they cannot influence the political process. Those who appear to find it very difficult may have the highest chance of disagreeing with the statement. As a mental note this may be worth checking with inferential statistics in another post.

#Satisfaction with the government

Last of all it is worth asking what are people's overall satisfaction with the government. This too is asked in the survey. 


```{r, include=FALSE, warning=FALSE}

# Very basic bar graph
p <- ggplot(data=poli, aes(x=stfgov))

```

```{r, echo=FALSE, warning=FALSE}

p + geom_bar(stat = "count",position=position_dodge())+
  labs(title = "Satisfaction in the Government",
         subtitle = "Continuous variable scoring satifaction out of 10",
         caption = "Source: European Social Survey 2016, Russia only") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) 
```

Regarding overall satisfaction with the government, results appear clustered in the middle. A group of results appear to cluster at 0, 5 (the middle), and 10 (the highest value). This is a curious result since such a high portion of respondents feel like they have no say or influence on the outcome. 

Again, it is worth asking of those who are comfortable and those who are struggling feel equally satisfied with the government.

```{r, include=FALSE}
opp_income <- poli %>%
   group_by(gndr, hincfel) %>%
   summarise(satisfied = mean(stfgov, na.rm = TRUE)) %>%
   filter(!is.na(satisfied)) %>%
   filter(!is.na(gndr)) %>%
   filter(!is.na(hincfel)) %>%
   ungroup()

```

The resulting graph is shown below.

```{r, echo=FALSE, warning=FALSE}


 ggplot(opp_income, aes(satisfied, hincfel))+
   geom_line(aes(group = hincfel)) +
   geom_point(aes(color = gndr), size = 1.5) +
   scale_x_continuous(limits = c(0, 8)) +
   ylim("comfortable", "coping","difficult","very difficult")+
  labs(title = "Satisfaction in the Government",
         subtitle = "Dot plot splitting satisfaction with \ngovernment by satisfaction with income",
         caption = "Source: European Social Survey 2016, Russia only")+
  theme_minimal() +
  theme(plot.title = element_text(size = 13, margin = margin(b = 12)),
        plot.subtitle = element_text(size = 12, color = "darkslategrey", margin = margin(b = 25)),
        plot.caption = element_text(size = 11, margin = margin(t = 10), color = "grey70", hjust = 0)) 
```

The output suggests differences between income groups are extremely minor. I added gender to the graph, mostly out of habit, but generally the categories seem roughly equally satisfied with the government. There is minor evidence that those who struggle are the least satisfied. However, this is not likely to be statistically significant, as this average lies close to the centre (between 3.5 and 4).

Overall it seems that satisfaction with government sits within a feeling of low influence or low impact. The ESS measures trust in social institutions, which I explore in another post. For the moment it seems there are some mixed results in the how Russians feel about politics. 


