---
title: 'Real Incomes and Income Growth: Part 2 Group Differences and Inequality'
author: "Ivan Privalko"
date: '2022-03-24'
slug: real-incomes-and-income-growth-part-2-group-differences-and-inequality
categories: inequality
tags: []
---


Building on the previous post, I want to look at group differences in trends tied to income over time.
The previous post showed how incomes rose after the year 2000 and how this growth declined in later years, possibly due to recession, maturity, or sanctions. 
These trend lines took the average rates for a given year and did not consider the change in inequality between groups over time.
In this post, I want to focus more specifically on income, with the idea that I will focus on wages in a future post. 

# Average income trends in quintiles

One way to consider this growth, is to look at quintile differences in real income trends over time, comparing the five groups of average ordered income for each year. 
This is shown in the plot below.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(data.table)
library(tidyverse)


load(file = "~/Projects/Blog/Sandbox/ru2.Rdata")



ggplot(ru2, aes(y=mean_income,
                x=year,
                group=income_quintile))+
  geom_point(aes(colour=income_quintile)) + 
  geom_line(aes(colour=income_quintile)) +
  scale_y_log10(labels=scales::comma) +
  theme_minimal()+
  labs(title="Trends in real income using 2020 prices by quintile (1994-2020)",
       subtitle = "Adjusting for inflation and accounting for denominational change, \nincome gaps have remained persistent over time",
       x='Year',
       y='Rubles (2020 Prices CPI adjusted)',
       caption = "Source: RLMS 1991-2020 \nAnalysis: Ivan Privalko") +
  scale_colour_discrete(guide = guide_legend(reverse = T),
                        name = "Income quintile", labels = c("Lowest", 
                                                            "Second", 
                                                            "Third",
                                                            "Fourth",
                                                            "Highest")) +
  theme(
    plot.title = element_text(hjust=0.5),
    plot.subtitle = element_text(hjust=0.5))
```
Here, the highest income quintile earned far above the average value.
Over time, we see that differences between categories widen and narrow, but generally the results show that income increased for most respondents between 2000 and 2010.
After 2010, incomes begin to plateau and even decline for certain groups.

# Income differences between percentiles as they are

Before considering these, we will briefly show the average earnings of each percentile for 2020 which are shown in 2020 rubles. 
This is shown in the figure below.
The 3rd percentile, on average, earns just 10,221 rubles per month. 
[We know from recent updates and policy reviews that the highest possible unemployment benefit was raised from 8,000 Russian rubles to 12,130 Russian rubles](https://www.statista.com/statistics/1291959/unemployment-benefits-russia/#:~:text=Changes%20in%20unemployment%20benefits%20due%20to%20COVID%2D19%20in%20Russia&text=In%20April%202020%2C%20the%20highest,increased%20to%204%2C500%20Russian%20rubles.), so some of these respondents are likely earning unemployment benefit at most. 
Further, the minimum possible unemployment benefit was increased to 4,500 Russian rubles.

This is in line with the findings presented here, the lowest paid individuals are likely receiving income from benefits and support like unemployment.
Given that the highest possible payment is higher than the amount that we consider here, this suggests we are capturing those on unemployment benefits well. 
Further, we note that the 10th percentile earns 15,000 rubles a month according to the data, which is useful for calculating the p90/p10 ratio.

Regarding top earners, we see that average earnings in the 97th percentile are 90,000 rubles per month, while the 90th percentile earns 60,000 rubles per month.
Once again, these earnings are high and constitute a p90/p10 ratio of 4, which is above many European countries.
For some perspective [the United States has a p90/p10 estimate of 6.3 while Japan has an estimate of 6.1](https://internationalcomparisons.org/social/income-distribution/); at the same time, Denmark (3.0), France(3.6), and Germany (3.6) have significantly lower estimates between these percentiles.

The estimate presented here brings Russia closer to Canada (4.0) and Australia (4.3) in terms of income inequality, which is closer aligned to many liberal market economies, than more egalitarian states like Finland, with whom Russia shares a border.


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

load(file = "~/Projects/Blog/Sandbox/percent.Rdata")


percent %>% 
  filter(year==2020,
         percentile <98 & percentile > 2) %>% 
  ggplot(aes(x=percentile, 
             y= income,
             group=percentile))+
  geom_col()+
  scale_y_continuous(labels=scales::comma) +
  geom_hline(yintercept = 36061.22)+
  geom_text(aes(30, 36061.22,label = "Mean income, 36,061", vjust = -1))+
  theme_minimal()+
  labs(title="Average earnings by percentile for 2020",
       subtitle = "Adjusting for inflation and accounting for denominational change. \nWe drop the bottom 2 and top 2 percentiles from our analysis",
       x='Income percentile',
       y='Rubles (2020 Prices CPI adjusted)',
       caption = "Source: RLMS 1991-2020 \nAnalysis: Ivan Privalko")+
  theme(
    plot.title = element_text(hjust=0.5),
    plot.subtitle = element_text(hjust=0.5))

```


We further explore this ratio later, for the moment it is worth noting the high rates of inequality between percentiles in 2020 data. 
We also want to consider how these figures looked in 2010, when Russian growth began to slow.
We find that earnings were similar in 2012, but that very high earners reported lower incomes in 2020 when compared to 2012. 
The bottom 3rd percentile earned just 8,200 rubles per month, while the 97th percentile earned 98,300 rubles per month. 

In this way we note a fall in some of the very top incomes between 2012 and 2020, we now turn to these falls for top earners.


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p1")

p1
```

In order to compare change over time, we can consider changes for percentiles between two time points. 
Given the extreme volatility of the Russian economy in the 1990's and the subsequent shocks, it is worth considering time points both overall and separately.
The graph below looks at the rate of change of each percentile for several time points (for example 2020 versus 2012).
The first graph considers four sets of time points; 2020 vs 2012, 2012 vs 2007, 2007 vs 2000, and 2000 vs 1995.
The y-axis compares the size of the change between the two time points. 
For example, for the 3rd percentile, income grew 4 times greater between 2000 and 2007; for the 97th percentile income grew almost 1.5 times greater during the same period. 


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p2")

p2
```

In two periods income grew for all percentiles, between 2000 and 2007, each percentile experienced an increase in income, though very low earners saw the greatest increase in income. 
Between 2007 and 2012, there was more modest growth but crucially, the rate of growth was equal across the income distribution. 
Between 2012 and 2020, income only gre for very poor respondents, remains in place for the middle class, and fell slightly for top earners. 
Finally, between 1995 and 2000, incomes fell gradually during a time of economic upheaval. 


Looking closely at the 2007-2012 we see growth of roughly 15%-60% depending on the income percentile, and some decline when looking at the results 2012-2020. 
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p4")

p4
```


Looking at the older results, we see a decline in growth 1995-2000, followed by strong growth especially for low earners between 2000 and 2007. 


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p3")

p3
```

# Percentile Ratios

The results of each graph suggest that growth and development has benefited those on lower incomes more than it has benefited those on higher incomes. 
In short this suggests that income inequality decreased quickly after the year 2000, slowly after 2007, and plateaued to minor closures of the gap from 2012.  
We can check this using the p90:p10 ratio mentioned earlier, which tracks the ratio between the 90th percentile income and the 10th percentile income.
We can also build on this, adding additional ratios like the p50:p10, and the p90:p50.
This is shown in the graph below.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p5")

p5
```

Here we see a gradual fall in inequality from 1996, with the 1990's capturing extreme differences of 10:1, the 2000's capturing lower levels, and the 2010's capturing a range of 6:1 and eventually 4:1.
However, this is only relevant for the extreme groups of the p90:p10 ratio.
The other ratios are more muted and typically hover around the ratio of 3:1, although here too there has been a decline in inequality. 

In general, Russia's p90:p10 ratio is on par with the p90:p10 ratio captured by the [OECD for 2019](https://www.oecd-ilsddsvibrary.org/docserver/b00d8f0e-en.pdf?expires=1648641701&id=id&accname=guest&checksum=F9236C6D10137E011AF86550924AA04A), which has a ratio of 4.6:1.
Further, the same statistics record a p50:p10 ratio of 2.2 for Russia, where we record a measure of 2 for the same year.


# Top decile share of income aan other decile shares

Further, we will consider the share of the top decile among Russia's total income. 
As before this measure is tracked from 1994.
This is listed below.
The Figure shows that the top decile's share of total income has fallen slightly, but that the change has been less extreme than the percentile share differences recorded above.


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p6")

p6
```


Finally, we can consider share of total income given to all quintiles since 1994. 
We see that the top earners' share of total income has gradually declined as other groups have grown. 
However, despite this, the top earners also have the largest share of total Russian income, almost 20%. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
load(file = "~/Projects/Blog/Sandbox/p7")

p7
```

In general we see that the lower income groups have benefited from economic growth and development, but that Russia still has high levels of inequality. 
I want to turn back to these measures for an analysis of labour income and wages, as well as recording differences between groups like educational qualifications and occupational groups. 
