---
title: Working with ESS data in R
author: Ivan
date: '2020-04-01'
slug: working-with-ess-data-in-r
categories: []
tags: [ESS]
---


I came across a good step by step guide for using [European Social Survey in R](http://dimiter.eu/Visualizations_files/ESS/Visualizing_ESS_data.html). 
I had some issues with the code, and in the end I found that `tidyverse` has some nice alternatives to the code preseneted, like the stuff on weighting, but in general this was a really interesting post.
I want to walk through some of my workflow here.
We can start by loading some of the ESS data for the Russian Federation.

```{r, echo=TRUE, results='hide', warning=FALSE, error=FALSE,message=FALSE}
library(tidyverse)
library(essurvey) # install from CRAN first
set_email("privalki@tcd.ie") # set your registered email
ru8 <- import_country(country = "Russian Federation",
               rounds = c(8),
               format = 'spss')


```

This is really easy with the new `essurvey` package which contains ESS data for all countries and all rounds considered.
Importantly, I originally left the format option blank, and this lead the format to be loaded as Stata. 
I am not exactly sure why but this led me to problems later on in the analysis, so I changed it. 

We can then explore some of the measures and controls using functions from `labelled`. Here I want to look at measures which reference income. 

```{r, echo=TRUE, results='hide', warning=FALSE, error=FALSE,message=FALSE}
library(labelled)
# explore and label -------------------------------------------------------

labelled::look_for(ru8, 'income')
attributes(ru8$dfincac)
summary(ru8$dweight) 
attributes(ru8$hinctnta)
summary(ru8$dweight) 
```



Before we visualise, we can use the `to_factor()` command to create a factor with the appropriate labels.
The code looks like this.


```{r, echo=TRUE, results='hide', warning=FALSE, error=FALSE,message=FALSE}
# recode ------------------------------------------------------------------

ru8$ineq <- to_factor(ru8$dfincac, drop_unused_labels=TRUE, ordered=TRUE)
table(ru8$ineq)


ru8$incdecile <- to_factor(ru8$hinctnta, drop_unused_labels=TRUE, ordered=TRUE)
table(ru8$incdecile)



```



We can also simplify the income agreement measure into a simpler measure, I'm just using it arbitrarily as an example for now.

```{r, echo=TRUE, results='hide', warning=FALSE, error=FALSE,message=FALSE}
# recode factor -----------------------------------------------------------

ru8 <- ru8 %>% 
  mutate(ineq = fct_recode(ineq, 
                    "agree" ="Agree strongly", 
                    "agree" = "Agree",
                    "disagree" = "Disagree",
                    "disagree" = "Disagree strongly")) 
```


We can then turn to `srvyr` to include survey weights in our results, which are slightly cumbersome, but will lead to weighted output.

```{r, echo=TRUE, results='hide', warning=FALSE, error=FALSE,message=FALSE}
# weighted data -----------------------------------------------------------
library(srvyr)
ru8_srvyr <- ru8 %>% 
  filter(!is.na(ineq),
         !is.na(incdecile)) %>% 
  as_survey_design(ids = 1,
                   weight = pspwght)


out <- ru8_srvyr %>%
  group_by(incdecile, ineq) %>% 
  summarize(proportion = survey_mean(),
            total = survey_total())

```


Having saved the weighted results of the proportion and survey total of the measure, we can then turn to graphing the results using `ggplot()`


```{r, echo=FALSE, results='hide', warning=FALSE, error=FALSE,message=FALSE}

out %>% 
  ggplot(aes(x=incdecile,
             y=proportion,
             fill=ineq))+
  geom_col()+
  coord_flip()+
  theme(legend.position="bottom")+
  labs(title = "When asked whether large differences \nin income are acceptable to reward talents and \nefforts, most disagree.",
       subtitle = "Even when we split the outcome by income decile, \nmost disagree.",
       y="Proportion of income decile",
       x="Income decile persone belongs to",
       fill="",
       caption = "Source: European Social Survey. \nPlot: @privlko")


```

There are still several things I want to try here, but the main point is that ESS data can quickly be ported into R and quickly analysed using weights with `srvyr`.








